A Novel Solution for Resonant Scattering Using Self-Consistent Boundary Conditions

### 2022-05-05 ms.tex
Final changes from Thursday meeting

\documentclass[linenumbers]{aastex63}
\usepackage{amsmath}
\newcommand{\be}{\begin{eqnarray}}
\newcommand{\ee}{\end{eqnarray}}
\def\la{\mathbin{\lower 3pt\hbox
      {$\rlap{\raise 5pt\hbox{$\char'074$}}\mathchar"7218$}}}
\def\ga{\mathbin{\lower 3pt\hbox
      {$\rlap{\raise 5pt\hbox{$\char'076$}}\mathchar"7218$}}} %> or of order
\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\grad}{\mathbf{\nabla}}
\newcommand\lya{Ly$\alpha$\ }

\newcommand{\todo}{\textbf{TODO: }}

\received{}
\revised{}
\accepted{\today}
\submitjournal{APJ}

@@

\title{A Novel Solution for Resonant Scattering Using Self-Consistent Boundary Conditions}
% the paper formerly known as Resonant Scattering in a Uniform Sphere with Large Optical Depth

@@

Given the abundance of hydrogen in the universe, \textbf{the Lyman $\alpha$ (Ly$\alpha$) line is an important component of radiation fields in a wide range of astrophysical settings. \lya radiation transport is an active area of research in the study of planets, stars, galaxies, and cosmology \citep{2019SAAS...46....1D}. An example application motivating our work is the role of \lya in planetary atmospheres.} The outer layers of the atmosphere are central to a planet's evolution, since they can shelter the lower atmosphere from high energy radiation as well as regulate the escape of gas into space.  There are two sources of Ly$\alpha$: the star, and recombinations in the planet's atmosphere. \lya may ionize atoms and dissociate molecules, as well as exert pressure forces that drive an outflow \textbf{ \citep{2018A&A...620A.147B}}. \lya can also excite H atoms to the 2p state, creating a population of Balmer-line absorbers that can be observed via transmission spectroscopy \citep{2017ApJ...851..150H}. Due to the low gas densities in the upper atmosphere, collisional de-excitation and broadening are of secondary importance and \lya may undergo ``resonant scattering''.

Hubble Space Telescope (HST) observations with the STIS have found large \lya transit depths around a handful of exoplanets \citep{2003Natur.422..143V, 2012A&A...543L...4L, 2012A&A...547A..18E, 2015Natur.522..459E,  2017A&A...597A..26B, 2017A&A...599L...3B, 2017A&A...602A.106B, 2018A&A...620A.147B, 2019AJ....158...50W, 2019EPSC...13.1928L, 2020ApJ...888L..21G,2021arXiv210309864B}. These observations have revealed a population of atoms extending out to distances of order a few planetary radii or more for several planets around bright, nearby stars, motivating a study of the physics of \lya interactions with the H atom population. The transition from the atomic to the molecular layer in these hot upper atmospheres may take place at pressures of order ${\sim}\ 10\ \mu$bar (see discussion in \citet{2017ApJ...851..150H} for details). This suggests the presence of a thick layer of atomic H which can have a line center optical depth of $\tau_0\ {\sim}\ 10^8$. A careful treatment of resonant scattering is necessary in order to construct accurate models of H atom excitation, heating, and radiative forces.

@@

\textbf{The motivation for including time-dependence in the transfer equation is to characterize the distribution of photon escape times, which is needed to calculate the radiation moments in the Monte Carlo simulation. Additionally, steady-state solutions to this problem are not always sufficient to describe all the physics of \lya transport. Time-variable, optically-thick environments necessitate a time-dependent solution to include the dynamic effects of \lya transfer. These include the optical afterglow of gamma-ray bursts \citep{2010ApJ...716..604R} and \lya sources redshifted by cosmological expansion %, which may reveal information about their neutral hydrogen halos and the intergalactic medium 
\citep{2011MNRAS.418..853X}, among others.}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{xinit.pdf}
    \caption{$P_{\rm bc}(x) = 16\pi^2R^2H_{\rm bc}(R, x)\Delta/L$ (Equation (\ref{eq:prob_spectrum}) with $H \rightarrow H_{\rm bc}$) vs. frequency $x$. %\textbf{Each panel shows $\sigma_s$ further from line center, labeled by $\sigma_s=0, \tau_0, 2\tau_0$. The solution is shown at a range of line center optical depths between $\tau_0=10^5$ and $10^9$ with solid overlapping lines. The thickness of the lines decreases with increasing optical depth to more clearly show where they overlap. The location of the core-wing boundary $x_{cw}$ for each $\tau_0$ is shown with dashed vertical lines which match the color of the corresponding solution}. The axes are scaled to show that the size of the correction factor agrees with the predicted $(a\tau_0)^{-1/3}$ scaling.
    \textbf{Each panel shows $\sigma_s$ further from line center, labeled by $\sigma_s=0, \tau_0, 2\tau_0$. The solution is shown at a range of line center optical depths between $\tau_0=10^5$ and $10^9$ with solid lines. The location of the core-wing boundary $x_{cw}$ for each $\tau_0$ is shown with dashed vertical lines which match the color of the corresponding solution}. The axes are scaled to show that the size of the correction factor agrees with the predicted $(a\tau_0)^{-1/3}$ scaling.}
    \label{fig:xinit}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics{Jsoln_n1_m100.pdf}
    \caption{Eigenfunctions $J_{nm}(x)$ for the lowest-order spatial eigenmode $n=1$, and $m=1, ..., 100$ with $x_s=0$ and $\tau_0=10^7$. \textbf{The scale of the $J_{nm}(\sigma)$ are set by the factor $E/(kR^3)$, which here is ${\sim}10^{-37}$ in units of ergs/(cm$^2$ Hz).}}
    \label{fig:jsoln}
\end{figure}

@@

The eigenfunction's oscillatory forms have varying amplitudes which sum in Equation (\ref{eq:Jrsigmat}) to create the final form of the mean intensity. The largest contribution at late times always comes from the ($n=1, m=1$) lowest-order eigenfunction. Figure \ref{fig:jsoln} shows a set of eigenfunctions $J_{nm}(\sigma)$ to illustrate their relative scales for different $m$ at a fixed spatial eigenmode $n$. The overall scale of the $J_{nm}(\sigma)$ are set by the factor $E/(kR^3)$ with $E$ arbitrarily set to 1. For H atoms with $T=10^4$ K and $\tau_0=10^7$, an eigenfunction has typical size ${\sim} E a / \left(R^2 \Delta \right) = 10^{-37}$ in units of specific mean intensity times time. Additional terms add smaller-magnitude, faster-oscillating components that lead to higher accuracy upon summation with the lower-order terms in Equation (\ref{eq:Jrsigmat}). The oscillations of various modes must cancel in the Fourier sum, so many modes $m$ and $n$ are required for convergence to the solution. 
\begin{figure}
    \centering
    \includegraphics[width=1.0\textwidth]{gamma_nm.pdf}
    \caption{Decay time vs. mode number. Overlapping solid lines are plotted for each $n$. The thickness of the lines decrease with $n$ in order to clearly show where they are in agreement. These numerically-obtained resonant frequencies are compared with the analytic expression in Equation (\ref{eq:gamma_nm}), shown as a dashed line. The parameters for the calculation are $\tau_0=10^7$ and $x_s=0$. The value of $n$ is given by the color bar on the right-hand side.}
    \label{fig:gamma_nm}
\end{figure}
The values of the $\gamma_{nm}$ can be described approximately with Equation (\ref{eq:gamma_nm}). Their values depend on $m$, $n$, and other physical parameters according to 
\be \label{eq:gamma_nm}
\gamma_{nm} = 2^{-1/3} \pi^{13/6} n^{4/3}\left(m-\frac{7}{8}\right)^{2/3}\frac{c}{R}(a\tau_0)^{-1/3}
\ee
as shown \textbf{by setting the denominator of Equation \ref{eq:a3/a2} to zero in WKB approximation of} Appendix \ref{app:wkb}. The power law in $m$ is weak, requiring up to $m=1000$ to reduce the scale of $\gamma_{nm}^{\ \ -1}$ by two orders of magnitude. When sweeping through to find resonances, Equation (\ref{eq:gamma_nm}) is used to set the scale of the sweep points $\gamma_j$ to ensure no $\gamma_{nm}$ are missed. The close agreement with the analytic expression shown in Figure \ref{fig:gamma_nm} indicate that the numerical solutions are accurate.

@@

\software{\texttt{numpy} \citep{2020NumPy-Array}, \texttt{scipy} \citep{2020SciPy-NMeth}}

>>>>

\documentclass[linenumbers]{aastex63}
\usepackage{amsmath}
\newcommand{\be}{\begin{eqnarray}}
\newcommand{\ee}{\end{eqnarray}}
\def\la{\mathbin{\lower 3pt\hbox
      {$\rlap{\raise 5pt\hbox{$\char'074$}}\mathchar"7218$}}}
\def\ga{\mathbin{\lower 3pt\hbox
      {$\rlap{\raise 5pt\hbox{$\char'076$}}\mathchar"7218$}}}
\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\grad}{\mathbf{\nabla}}
\newcommand\lya{Ly$\alpha$\ }

\submitjournal{ApJ}

@@

\title{A Novel Solution for Resonant Scattering Using Self-Consistent Boundary Conditions}

@@

Given the abundance of hydrogen in the universe, \textbf{the Lyman $\alpha$ (Ly$\alpha$) line is an important component of radiation fields in a wide range of astrophysical settings. \lya radiation transport is an active area of research in the study of planets, stars, galaxies, and cosmology \citep{2019SAAS...46....1D}. An example application motivating our work is the role of \lya in planetary atmospheres.} The outer layers of the atmosphere are central to a planet's evolution, since they can shelter the lower atmosphere from high energy radiation as well as regulate the escape of gas into space.  There are two sources of Ly$\alpha$: the star, and recombinations in the planet's atmosphere. \lya may ionize atoms and dissociate molecules, as well as exert pressure forces that drive an outflow \textbf{ \citep{2018A&A...620A.147B}}. \lya can also excite H atoms to the 2p state, creating a population of Balmer-line absorbers that can be observed via transmission spectroscopy \citep{2017ApJ...851..150H, 2021ApJ...907L..47Y}. Due to the low gas densities in the upper atmosphere, collisional de-excitation and broadening are of secondary importance and \lya may undergo ``resonant scattering''.

Hubble Space Telescope (HST) observations with the STIS have found large \lya transit depths around a handful of exoplanets \citep{2003Natur.422..143V, 2012A&A...543L...4L, 2012A&A...547A..18E, 2015Natur.522..459E,  2017A&A...597A..26B, 2017A&A...599L...3B, 2017A&A...602A.106B, 2018A&A...620A.147B, 2019AJ....158...50W, 2019EPSC...13.1928L, 2020ApJ...888L..21G,2021arXiv210309864B}. These observations have revealed a population of atoms extending out to distances of order a few planetary radii or more for several planets around bright, nearby stars, motivating a study of the physics of \lya interactions with the H atom population. The transition from the atomic to the molecular layer in these hot upper atmospheres may take place at pressures of order ${\sim}\ 10\ \mu$bar (see discussion in \citealt{2017ApJ...851..150H} for details). This suggests the presence of a thick layer of atomic H which can have a line center optical depth of $\tau_0\ {\sim}\ 10^8$. A careful treatment of resonant scattering is necessary in order to construct accurate models of H atom excitation, heating, and radiative forces.

@@

\textbf{The motivation for including time-dependence in the transfer equation is to characterize the distribution of photon escape times, which is needed to calculate the radiation moments in the Monte Carlo simulation. Additionally, steady-state solutions to this problem are not always sufficient to describe all the physics of \lya transport. Time-variable, optically-thick environments necessitate a time-dependent solution to include the dynamic effects of \lya transfer. These include the optical afterglow of gamma-ray bursts \citep{2010ApJ...716..604R} and \lya sources redshifted by cosmological expansion
\citep{2011MNRAS.418..853X}, among others.}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{xinit.pdf}
    \caption{$P_{\rm bc}(x) = 16\pi^2R^2H_{\rm bc}(R, x)\Delta/L$ (Equation (\ref{eq:prob_spectrum}) with $H \rightarrow H_{\rm bc}$) vs. frequency $x$. \textbf{Each panel shows $\sigma_s$ further from line center, labeled by $\sigma_s=0, \tau_0, 2\tau_0$. The solution is shown at a range of line center optical depths between $\tau_0=10^5$ and $10^9$ with solid lines. The location of the core-wing boundary $x_{cw}$ for each $\tau_0$ is shown with dashed vertical lines which match the color of the corresponding solution}. The axes are scaled to show that the size of the correction factor agrees with the predicted $(a\tau_0)^{-1/3}$ scaling.}
    \label{fig:xinit}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics{Jsoln_n1_m100.pdf}
    \caption{Eigenfunctions $J_{nm}(x)$ for the lowest-order spatial eigenmode $n=1$, and $m=1, ..., 100$ with $x_s=0$ and $\tau_0=10^7$. \textbf{The scale of the $J_{nm}(\sigma)$ are set by the factor $E/(kR^3)$, which here is ${\sim}10^{-37}$ in ergs/(cm$^2$ Hz).}}
    \label{fig:jsoln}
\end{figure}

@@

The eigenfunction's oscillatory forms have varying amplitudes which sum in Equation (\ref{eq:Jrsigmat}) to create the final form of the mean intensity. The largest contribution at late times always comes from the ($n=1, m=1$) lowest-order eigenfunction. Figure \ref{fig:jsoln} shows a set of eigenfunctions $J_{nm}(\sigma)$ to illustrate their relative scales for different $m$ at a fixed spatial eigenmode $n$. The overall scale of the $J_{nm}(\sigma)$ are set by the factor $E/(kR^3)$ with $E$ arbitrarily set to 1. For H atoms with $T=10^4$ K and $\tau_0=10^7$, an eigenfunction has typical size ${\sim} E a / \left(R^2 \Delta \right) = 10^{-37}$ in units of specific mean intensity times time. Additional terms add smaller-magnitude, faster-oscillating components that lead to higher accuracy upon summation with the lower-order terms in Equation (\ref{eq:Jrsigmat}). The oscillations of various modes must cancel in the Fourier sum, so many modes $m$ and $n$ are required for convergence to the solution. 
\begin{figure}
    \centering
    \includegraphics[width=1.0\textwidth]{gamma_nm.pdf}
    \caption{Dimensionless decay time vs. mode number. Overlapping solid lines are plotted for each $n$. The thickness of the lines decrease with $n$ in order to clearly show where they are in agreement. These numerically-obtained resonant frequencies are compared with the analytic expression in Equation (\ref{eq:gamma_nm}), shown as a dashed line. The parameters for the calculation are $\tau_0=10^7$ and $x_s=0$. The value of $n$ is given by the color bar on the right-hand side.}
    \label{fig:gamma_nm}
\end{figure}
The values of the $\gamma_{nm}$ can be described approximately with Equation (\ref{eq:gamma_nm}). Their values depend on $m$, $n$, and other physical parameters according to 
\be \label{eq:gamma_nm}
\gamma_{nm} = 2^{-1/3} \pi^{13/6} n^{4/3}\left(m-\frac{7}{8}\right)^{2/3}\frac{c}{R}(a\tau_0)^{-1/3}
\ee
as shown \textbf{by setting the denominator of Equation \ref{eq:a3/a2} to zero in WKB approximation of} Appendix \ref{app:wkb}. The power law in $m$ is weak, requiring up to $m=1000$ to reduce the scale of $\gamma_{nm}^{\ \ -1}$ by two orders of magnitude. When sweeping through to find resonances, Equation (\ref{eq:gamma_nm}) is used to set the scale of the sweep points $\gamma_j$ to ensure no $\gamma_{nm}$ are missed. The close agreement with the analytic expression shown in Figure \ref{fig:gamma_nm} indicate that the numerical solutions are accurate.

@@

\software{\texttt{numpy} \citep{2020NumPy-Array}, \texttt{scipy} \citep{2020SciPy-NMeth}, Coblis - Color Blindness Simulator (\href{https://www.color-blindness.com/coblis-color-blindness-simulator/}{color-blindness.com})}}
